From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-21 dictionary.tex
Fixed definition of QCS

Let $K$ be a local field with finite residue field $\Fq$; we will now denote the group $\Weil{}$ by $\Weil{\Fq}$.
We continue to assume that $\ell$ is invertible in $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.  This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.
We write $\mathcal{N}$ for the category of Neron models that arise in this way; in particular, $\mathcal{N}$ is a full subcategory of the category of smooth commutative group schemes over $R$.

@@

 \begin{remark}
This paper shows that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a Neron model for an algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed. 
This seems unlikely, since the geometric etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$, though limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

Let $X \in \mathcal{N}$ be as above.
Set $S = \Spec{R}$ and $S_n = \Spec{R_n}$;
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.

For every non-negative integer $n$, $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$
with component group scheme $\pi_0(X)_{\Fq} \ceq \pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
with component group scheme $\pi_0(X)_{\Fq}$
and $\Fq$-rational points $X(K).$
However, $\Gr_R(X)$ is not locally of finite type and therefore not smooth.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(X)$ induces a functor
\[
\varrho^R_n(X)^* : \CS(\Gr^R_n(X)) \to \CS(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\CS(\Gr^R_n(X))$ as a subcategory of $\CS(\Gr_R(X))$.

We now define a category which will geometrize quasicharacters of $X(K)$.
We begin by considering the rigid monoidal category $\CS(G)$ defined in Section~\ref{ssec:category} without insisting that the commutative group $\Fq$-scheme $G$ is smooth, so that we can replace it with $\Gr_R(X)$ for smooth group $R$-scheme $X$.

\begin{definition}\label{def:QCS}
Let $X$ be a smooth commutative group scheme over $R$, with $R$ as above.
A {\it quasicharacter sheaf for $X$} is an object of
the following rigid monoidal subcategory of $\CS(\Gr_R(X))$, denoted by $\QCS(X)$:
\begin{enumerate}
\item
objects in $\QCS(X)$ are the $\ell$-adic sheaves $\varrho^R_n(X)^*\cs{L}$, for $n\in \NN$ and $\cs{L} \in \CS(\Gr^R_n(X))$; 
\item
morphisms $\varrho^R_n(X)^*\cs{L} \to \varrho^R_m(X)^*\cs{L}'$ in $\QCS(X)$ are those morphisms in $\CS(\Gr_R(X))$ which take the form $\varrho^R_m(X)^*\rho$ for $\rho \in \Hom(\varrho^R_{n\leq m}(X)^*\cs{L},\cs{L}')$ when $n\leq m$, and $\varrho^R_n(X)^*\rho$ for $\rho \in \Hom(\cs{L},\varrho^R_{m\leq n}(X)^*\cs{L}')$ when $m\leq n$.
\end{enumerate}

\end{definition}

@@

\begin{remark}
We offer the following alternate construction of $\QCS(X)$, which does not make use of subcategories.
For objects, one considers triples 
\[
\cs{F} \ceq (n, \{\cs{F}_i\}_{n\leq i}, \{\alpha_{i \le j}\}_{n\le i \le j}),
\] where $n$
is a non-negative integer, each $\cs{F}_i$ is a character sheaf on $\Gr^R_i(X)$ and each 
\[
\alpha_{i \le j} : \cs{L}_j \to \varrho^R_{i \le j}(X)^* \cs{L}_i
\]
 is an isomorphism; here $\alpha_{i \le i}$ is the identity and the $\alpha_{i \le j}$ are compatible with each other.  
If $\cs{F} \ceq (n, \{\cs{F}_i\}, \{\alpha_{i \le j}\})$
and $\cs{F}' \ceq (m, \{\cs{F}'_i\}, \{\alpha'_{i \le j}\})$ are objects then $\Hom(\cs{F}, \cs{F}')$ is the set
of equivalence classes of pairs $(k, \{\beta_i\}_{k \le i})$, where $n,m \le k$ and the $\beta_i : \cs{F}_i \to \cs{F}'_i$ are
morphisms of character sheaves so that
\[
\begin{tikzcd}
\cs{F}_j \rar{\alpha_j} \dar{\beta_i} & \varrho^R_{i \le j}(X)^* \cs{F}_i \dar{f_{i \le j}^*\beta_j} \\
\cs{F}'_j \rar{\alpha_j} & \varrho^R_{i \le j}(X)^* \cs{L}'_i
\end{tikzcd}
\]
commutes for all $k\le i\le j$; we identify two such pairs $(k, \{\beta_i\})$ and $(l, \{\gamma_i\})$ if $\beta_i = \gamma_i$
for sufficiently large $i$.  Identities and composites are defined in the natural way.
\end{remark}

@@

@@

\begin{corollary}\label{cor:CS_tori}
Let $T$ be a Neron model for an algebraic torus over $K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{inf}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{res}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ preserve limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of $\TrFrob{T}$ is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{\Fq}} = \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}.
\] 
Thus, the middle vertical sequence is exact.
Since $T^0$ and $\pi_0(T)$ do not lie in $\mathcal{N}$, we cannot use Theorem~\ref{thm:CSXK} to determine the image and trace of Frobenius for these schemes. 
Instead, we observe that $T^0$ and $\pi_0(T)$ are smooth commutative group schemes over $R$, so Definition~\ref{def:QCS} gives meaning to categories $\QCS(T^0)$ and $\QCS(\pi_0(T))$, and, moreover, that $\pi_0(\Gr^R_n(T^0))= \pi_0(T^0)_{\Fq} = 1$ and $\pi_0(\Gr^R_n(\pi_0(T)) = \pi_0(T)_{\Fq}$ are both independent of $n$.
It follows that the vertical sequences through $\QCSiso{T^0}$ and $\QCSiso{\pi_0(T)}$ are exact by Theorem~\ref{thm:snake} and Definition~\ref{def:QCS}.
The diagram commutes by Lemma~\ref{lem:pullback}.
\end{proof}

>>>>

Let $K$ be a local field with finite residue field $\Fq$; in this section we denote the group $\Weil{}$ by $\Weil{\Fq}$.
We continue to assume that $\ell$ is invertible in $\Fq$. 
In this section we consider connected commutative algebraic groups over $K$ that admit a N\'eron model, by which we mean a locally finite type N\'eron model.
By \cite{bosch-lutkebohmert-reynaud:NeronModels}*{\S 10.2, Thm 2}, these are precisely the connected commutative algebraic groups over $K$ that contain no subgroup isomorphic to $\mathbb{G}_\text{a}$.
Write $\mathcal{N}_K$ for the full subcategory of the category of algebraic groups consisting of such objects.  This category is additive, and includes all algebraic tori over $K$, abelian varieties over $K$ and unipotent $K$-wound groups.
We write $\mathcal{N}$ for the category of Neron models that arise in this way; in particular, $\mathcal{N}$ is a full subcategory of the category of smooth commutative group schemes over $R$.

@@

%Let $X \in \mathcal{N}$ be as above.
Set $S = \Spec{R}$ and $S_n = \Spec{R_n}$;
note that $S_0 = \Spec{\Fq}$ is the special fibre of $S$.
Let $X$ be a smooth commutative group scheme over $S$.
%We now define a category which will geometrize quasicharacters of $X(K)$.
For every non-negative integer $n$, the Greenberg transform $\Gr^R_n(X)$ is a {\it smooth} commutative group scheme over $S_0$
%with component group scheme $\pi_0(X)_{\Fq} \ceq \pi_0(X) \times_S S_0$ and $\Fq$-rational points $X(R_n).$
%
The Greenberg transform $\Gr_R(X)$ of $X$ is a commutative group scheme over $\Fq$
%with component group scheme $\pi_0(X)_{\Fq}$
with $\Fq$-rational points $X(R)$.
%
The morphism of $\Fq$-schemes $\varrho^R_n(X) : \Gr_R(X) \to \Gr^R_n(X)$ induces a functor
\[
\varrho^R_n(X)^* : \CS(\Gr^R_n(X)) \to \CS(\Gr_R(X)),
\]
as in Lemma~\ref{lem:pullback}.
%This functor is faithful, so we may view $\CS(\Gr^R_n(X))$ as a subcategory of $\CS(\Gr_R(X))$.

\begin{definition}\label{def:QCS}
Let $X$ be a smooth group scheme over $R$.
A {\it quasicharacter sheaf} of $X$ is a triple 
\[
\cs{F} \ceq (n, \{\cs{F}_i\}_{n\leq i}, \{\alpha_{i \le j}\}_{n\le i \le j}),
\] 
where $n$
is a non-negative integer, each $\cs{F}_i$ is a character sheaf on $\Gr^R_i(X)$ and each 
\[
\alpha_{i \le j} : \cs{L}_j \to \varrho^R_{i \le j}(X)^* \cs{L}_i
\]
 is an isomorphism; here $\alpha_{i \le i}$ is the identity and the $\alpha_{i \le j}$ are compatible with each other.  
If $\cs{F} \ceq (n, \{\cs{F}_i\}, \{\alpha_{i \le j}\})$
and $\cs{F}' \ceq (m, \{\cs{F}'_i\}, \{\alpha'_{i \le j}\})$ are objects then $\Hom(\cs{F}, \cs{F}')$ is the set
of equivalence classes of pairs $(k, \{\beta_i\}_{k \le i})$, where $n,m \le k$ and the $\beta_i : \cs{F}_i \to \cs{F}'_i$ are
morphisms of character sheaves so that
\[
\begin{tikzcd}
\cs{F}_j \rar{\alpha_j} \dar{\beta_i} & \varrho^R_{i \le j}(X)^* \cs{F}_i \dar{f_{i \le j}^*\beta_j} \\
\cs{F}'_j \rar{\alpha_j} & \varrho^R_{i \le j}(X)^* \cs{L}'_i
\end{tikzcd}
\]
commutes for all $k\le i\le j$; we identify two such pairs $(k, \{\beta_i\})$ and $(l, \{\gamma_i\})$ if $\beta_i = \gamma_i$
for sufficiently large $i$.  Identities and composites are defined in the natural way.
Let $\QCS(X)$ denote the category of quasicharacter sheaves for $X$.
\end{definition}

@@

\begin{remark}
We offer the following alternate construction of $\QCS(X)$.
As above, let $X$ be a smooth group scheme over $R$.
Although $\Gr_R(X)$ is not locally of finite type and therefore not smooth,
let us consider the rigid monoidal category $\CS(\Gr_R(X))$ as defined in Section~\ref{ssec:category} without insisting that the commutative group $\Fq$-scheme $G$ is smooth.
%Let $X$ be a smooth commutative group scheme over $R$, with $R$ as above.
A quasicharacter sheaf for $X$ is an object of
the following rigid monoidal subcategory of $\CS(\Gr_R(X))$, denoted by $\QCS(X)$:
\begin{enumerate}
\item
objects in $\QCS(X)$ are the $\ell$-adic sheaves $\varrho^R_n(X)^*\cs{L}$, for $n\in \NN$ and $\cs{L} \in \CS(\Gr^R_n(X))$; 
\item
morphisms $\varrho^R_n(X)^*\cs{L} \to \varrho^R_m(X)^*\cs{L}'$ in $\QCS(X)$ are those morphisms in $\CS(\Gr_R(X))$ which take the form $\varrho^R_m(X)^*\rho$ for $\rho \in \Hom(\varrho^R_{n\leq m}(X)^*\cs{L},\cs{L}')$ when $n\leq m$, and $\varrho^R_n(X)^*\rho$ for $\rho \in \Hom(\cs{L},\varrho^R_{m\leq n}(X)^*\cs{L}')$ when $m\leq n$.
\end{enumerate}
\end{remark}

@@

 \begin{remark}
In Section~\ref{ssec:CS_tori} we see that the \'etale site on $\Gr_R(X)$ is rich enough to geometrize all quasicharacters of $X(K)$ as $\ell$-adic local systems on $\Gr_R(X)$, where $X$ is a Neron model for an algebraic torus or an abelian variety over a local field $K$. 
It is natural to ask if the \'etale site on the generic fibre $X_K$ would have sufficed. 
This seems unlikely, since the geometric etale fundamental group of $\Gm{K}$ is ${\hat \ZZ}$, though limited results in this direction were established in \cite{Cunningham-Kamgarpour} when $K = \mathbb{Q}_p$.
\end{remark}

@@

\begin{corollary}\label{cor:CS_tori}
Let $T$ be a Neron model for an algebraic torus over $K$.
The following is a commutative diagram of exact sequences.
\[
  \begin{tikzcd}[row sep=20, column sep=18]
{}  & 0 \arrow{d} & 0 \arrow{d} &  & \\ 
   & \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}  \arrow{d} & \arrow{d} \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}} & 0 \arrow{d} & \\
    0 \rar & \QCSiso{\pi_0(T)} \arrow{r}{\pi_0^*} \dar{\TrFrob{\pi_0(T)}}
    & \QCSiso{T} \arrow{r}{\iota_0^*}  \arrow{d}{\TrFrob{\Gr_R(T)}} & \QCSiso{T^0} \arrow{r} \arrow{d}{\TrFrob{\Gr_R(T)^0}} & 0\\
    0 \arrow{r} & \Hom(\pi_0(T)(\Fq), \EEx) \arrow{r}{\text{inf}} \arrow{d}{\iso}
    & \arrow{d}{\iso} \Hom_\text{}(T(K),\EEx) \arrow{r}{\text{res}} & \Hom_\text{}(T^0(R),\EEx) \arrow{r} \arrow{d}{\iso} & 0\\
 &  0  & 0 & 0 & 
  \end{tikzcd}
 \]
 \end{corollary}
\begin{proof} 
The horizontal sequence of groups coming from categories of quasicharacter sheaves is exact by Proposition~\ref{prop:middleexact}, together with the observation that the functors $\pi_0^*$ and $\iota^*$ preserve limits.
It is elementary that the horizontal sequence of quasicharacters is exact.
Accordingly, by Theorem~\ref{thm:CSXK}, the kernel of the  is $\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{}}$. 
By \cite{bitan:discriminant}*{Eq 3.1}, the special fibre of the component group scheme for $T$ is given by
\[
 \pi_0(T)_{\bFq} = X_*(T_K)_{\mathcal{I}_K},
\]
where $X_*(T_K)$ is the cocharacter lattice of $T_K$ and $\mathcal{I}_K$ is the inertia group for $K$.
Thus,
\[
\Hh^2(\pi_0(T)_{\bFq},\EEx)^{\Weil{\Fq}} = \Hh^2(X_*(T_K)_{\mathcal{I}_K},\EEx)^{\Weil{\Fq}}.
\] 
Thus, the middle vertical sequence is exact.
Since $T^0$ and $\pi_0(T)$ do not lie in $\mathcal{N}$, we cannot use Theorem~\ref{thm:CSXK} to determine the image and trace of Frobenius for these schemes. 
Instead, we observe that $T^0$ and $\pi_0(T)$ are smooth commutative group schemes over $R$, so Definition~\ref{def:QCS} gives meaning to categories $\QCS(T^0)$ and $\QCS(\pi_0(T))$, and, moreover, that $\pi_0(\Gr^R_n(T^0))= \pi_0(T^0)_{\Fq} = 1$ and $\pi_0(\Gr^R_n(\pi_0(T)) = \pi_0(T)_{\Fq}$ are both independent of $n$.
It follows that the vertical sequences through $\QCSiso{T^0}$ and $\QCSiso{\pi_0(T)}$ are exact by Theorem~\ref{thm:snake} and Definition~\ref{def:QCS}.
The diagram commutes by Lemma~\ref{lem:pullback}.
\end{proof}
